---
title: "Week 5 Computational Tools for Bioengineers"
author: "Bill Cresko"
format: 
  revealjs:
    footer: BioE_Comp_Tools_2025 - Knight Campus 
    transition: fade
    transition-speed: slow
editor: visual
---

```{r}
library(tidyverse)
library(gt)
library(readxl)
theme_set(theme_minimal())
```

## Goals for today

-   Using R Studio and Visual Studio Code (VSC) as script editors
-   R and Quarto Markdown
-   Vibe Coding with LLMs
-   Git and GitHub

## What is a Shell Script? {.smaller}

-   Text file containing Unix commands
-   Automates repetitive tasks
-   Makes complex workflows reproducible
-   Essential for bioinformatics pipelines
-   Can include:
    -   Variables and arrays
    -   Conditionals (if/then/else)
    -   Loops (for/while)
    -   Functions
    -   Command-line arguments

## Creating Your First Script {.smaller}

``` bash
#!/bin/bash
# This is a shebang - tells system which interpreter to use

# This is a comment - documents your code
echo "Hello, Bioengineering!"

# Set a variable
NAME="DNA Analysis"
echo "Running $NAME"

# Use command output as variable
DATE=$(date +%Y-%m-%d)
echo "Analysis date: $DATE"

# Simple calculation
COUNT=5
DOUBLE=$((COUNT * 2))
echo "Double of $COUNT is $DOUBLE"
```

Save as `first_script.sh` and run with `bash first_script.sh`

## Making Scripts Executable {.smaller}

``` bash
# Check current permissions
ls -l first_script.sh

# Make executable (add execute permission)
chmod +x first_script.sh

# Now you can run it directly
./first_script.sh
```

Best practice: Always use `chmod +x` for your scripts


# R statistical programming language

## Why use `R`? {.smaller}

-   Good general scripting tool for statistics and mathematics
-   Powerful and flexible and free
-   Runs on all computer platforms
-   New enhancements coming out all the time
-   Superb data management & graphics capabilities
-   Reproducibility - can keep your scripts to see exactly what was done
-   You can write your own functions
-   Lots of online help available
-   Can use a nice GUI front end such as `Rstudio`
-   Can embed your `R` analyses in dynamic, polished files using `Quarto Markdown`
-   Quarto Markdown can be reused for websites, papers, books, presentations...

## `R scripts` and `Quarto Markdown` {.smaller}

-   Often we want to write scripts that can just be run - `R scripts`
-   We can also embed code in Quarto Markdown files that provide more annotations
-   This improves interpertability and reproducibility
-   You can insert `R code chunks` into `Quarto Markdown` documents

## `R script` basics {.smaller}

-   A series of R commands that will be executed
\

-   Can add comments using hashtags `#`
\

-   Can have pipes (`|>`) to connect one step to the next
\


## BASICS of `R` {.flexbox .vcenter .smaller}

-   Commands can be submitted through the terminal, console or scripts
-   Notice on these slides I'm evaluating the code chunks and showing output
-   Also notice that `R` follows the normal priority of mathematical evaluation

\
\

```{r basic multiplication, echo=TRUE}
4*4
```

```{r more multiplication, echo=TRUE}
(4+3*2^2)
```

## Assigning Variables {.flexbox .vcenter}

-   A better way to do this is to assign variables
-   Variables are assigned values using the `<-` operator.
-   Variable names must begin with a letter, but other than that, just about anything goes.
-   Do keep in mind that `R` is case sensitive.

## Assigning Variables {.flexbox .vcenter}

```{r assigning variables, echo=TRUE, tidy=TRUE}

x <- 2
x*3
y <- x * 3
y-2

```

These do not work

```{r eval=FALSE, echo=TRUE}
3y <- 3
3*y <- 3
```

## Arithmetic operations on functions {.smaller}

-   Arithmetic operations can be performed easily on functions as well as numbers.
-   Try the following, and then your own.

```{r functions, eval = FALSE, echo = TRUE}
x+2
x^2
log(x)
```

-   Note that the last of these - `log` - is a built in function of `R`, and therefore the object of the function needs to be put in parentheses
-   These parentheses will be important, and we'll come back to them later when we add arguments after the object in the parentheses\
-   The outcome of calculations can be assigned to new variables as well, and the results can be checked using the 'print' command

## Arithmetic operations on functions {.smaller}

```{r, echo=TRUE}
y <- 67
print(y)

x <- 124
z <- (x*y)^2
print(z)
```

## STRINGS {.smaller}

-   Variables and operations can be performed on characters as well
-   Note that characters need to be set off by quotation marks to differentiate them from numbers
-   The `c` stands for `concatenate`
-   Note that we are using the same variable names as we did previously, which means that we're overwriting our previous assignment
-   A good rule of thumb is to use new names for each variable, and make them short but still descriptive

## STRINGS {.smaller}

```{r, echo=TRUE}
x <- "I Love"
print (x)
y <- "Bioengineering"
print (y)
z <- c(x,y)
print (z)
```

## VECTORS {.smaller}

-   In general `R` thinks in terms of vectors (a list of characters, factors or numerical values).
-   It will benefit any `R` user to try to write programs with that in mind, as it will simplify most things.
-   Vectors can be assigned directly using the `c()` function and then entering the exact values.

## VECTORS {.smaller}

```{r Example vectors, echo=TRUE}
x <- c(2,3,4,2,1,2,4,5,10,8,9)
print(x)
```

```{r Example vector multiplication, echo=TRUE}
x <- c(2,3,4,2,1,2,4,5,10,8,9)
y <- x^2
print(x)
print(y)
```

```{r Example vector plot, echo=TRUE}
plot(x,y)
```

## Basic Statistics {.smaller}

-   Many functions exist to operate on vectors.
-   Combine these with your previous variable to see what happens.
-   Also, try to find other functions (e.g. standard deviation).

```{r, eval = FALSE, echo = TRUE}
mean(x)
median(x)
var(x)
log(x)
ln(x)
sqrt(x)
sum(x)
length(x)
sample(x, replace = T)
```

-   Notice that the last function (`sample`) has an argument (`replace=T`)
-   Arguments simply modify or direct the function in some way
-   There are many arguments for each function, some of which are defaults

## Getting Help {.smaller}

-   Getting Help on any function is very easy - just type a question mark and the name of the function.
-   There are functions for just about anything within `R` and it is easy enough to write your own functions if none already exist to do what you want to do.
-   In general, function calls have a simple structure: a function name, a set of parentheses and an optional set of parameters to send to the function.
-   Help pages exist for all functions that, at a minimum, explain what parameters exist for the function.\
-   Help can be accessed a few ways - try them :

## Getting Help

```{r, eval = FALSE, echo = TRUE}
- help(mean)
- ?mean
- example(mean)
- help.search("mean")
- apropos("mean")
- args(mean)
```

## Creating vectors {.smaller}

-   Creating vector of new data by entering it by hand can be a drag
-   However, it is also very easy to use functions such as `seq` and `sample`
-   Try the examples below; can you figure out what the three arguments in the parentheses mean?
-   Try varying the arguments to see what happens.
-   Don't go too crazy with the last one or your computer might slow way down

## Creating vectors {.smaller}

```{r, echo = TRUE}
seq_1 <- seq(0.0, 10.0, by = 0.1)
print(seq_1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
print(seq_2)
```

## Creating vectors {.smaller}

```{r, echo = TRUE}
seq_square <- (seq_2)*(seq_2)
print(seq_square)
```

## Creating vectors {.smaller}

```{r, echo = TRUE}
seq_square_new <- (seq_2)^2
print(seq_square_new)
```

## Types of vectors of data {.smaller}

-   `int` stands for integers

-   `dbl` stands for doubles, or real numbers

-   `chr` stands for character vectors, or strings

-   `dttm` stands for date-times (a date + a time)

-   `lgl` stands for logical, vectors that contain only TRUE or FALSE

-   `fctr` stands for factors, which R uses to represent categorical variables with fixed possible values

-   `date` stands for dates

<!-- -->

-   Integer and double vectors are known collectively as numeric vectors.
-   In `R` numbers are doubles by default.

## Types of vectors of data {.smaller}

-   Logical vectors can take only three possible values:
    -   `FALSE`
    -   `TRUE`
    -   `NA` which is 'not available'.
-   Integers have one special value: NA, while doubles have four:
    -   `NA`
    -   `NaN` which is 'not a number'
    -   `Inf`
    -   `-Inf`\## Reading in Data Frames in R {.smaller}
-   A strength of `R` is being able to import data from an external source
-   Create the same table that you did above in a spreadsheet like Excel
-   Export it to comma separated and tab separated text files for importing into `R`.
-   The first will read in a comma-delimited file, whereas the second is a tab-delimited
-   In both cases the header and row.names arguments indicate that there is a header row and row label column
-   Note that the name of the file by itself will have R look in the CWD, whereas a full path can also be used

## Drawing samples from distributions {.smaller}

-   Here is a way to create your own data sets that are random samples.
-   Again, play around with the arguments in the parentheses to see what happens.

## Drawing samples from distributions {.smaller}

```{r Samples from distributions 1, out.width='100%', echo = TRUE}
x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
plot(x,y) 
```

## Drawing samples from distributions {.smaller}

```{r Samples from distributions 2, out.width='100%', echo = TRUE}
x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
plot(xy)
```

## Drawing samples from distributions {.smaller}

```{r Samples from distributions 3, out.width='100%', echo = TRUE}
x <- rnorm (10000, 0, 10)
y <- sample (1:10000, 10000, replace = T)
xy <- cbind(x,y)
hist(x)
```

## Drawing samples from distributions {.smaller}

-   You’ve probably figured out that y from the last example is drawing numbers with equal probability.
-   What if you want to draw from a distribution?
-   Again, play around with the arguments in the parentheses to see what happens.

## Drawing samples from distributions {out.width="50%"}

```{r, out.width='100%', echo = TRUE}
x <-rnorm(1000, 0, 100)
hist(x, xlim = c(-500,500))
curve(50000*dnorm(x, 0, 100), xlim = c(-500,500), add=TRUE, col='Red')
```

-   `dnorm()` generates the probability density, which can be plotted using the `curve()` function.
-   Note that is curve is added to the plot using `add=TRUE`

## Visualizing Data {.smaller}

-   So far you've been visualizing just the list of output numbers
-   Except for the last example where I snuck in a `hist` function.
-   You can also visualize all of the variables that you've created using the `plot` function (as well as a number of more sophisticated plotting functions).
-   Each of these is called a `high level` plotting function, which sets the stage
-   `Low level` plotting functions will tweak the plots and make them beautiful

## Visualizing Data {.smaller}

-   What do you think that each of the arguments means for the plot function?
-   A cool thing about `R` is that the options for the arguments make sense.
-   Try adjusting an argument and see if it works
-   Note- most people use the package `GGPlot2` for nice scientific graphics

## Visualizing Data {.smaller}

```{r, echo = TRUE}
seq_1 <- seq(0.0, 10.0, by = 0.1) 
plot (seq_1, xlab="space", ylab ="function of space", type = "p", col = "red")
```

## Putting plots in a single figure {.smaller}

-   On the next slide
-   The first line of the lower script tells R that you are going to create a composite figure that has two rows and two columns. Can you tell how?
-   Now, modify the code to add two more variables and add one more row of two panels.

```{r, echo=TRUE, outwidth="50%", out.height="20%"}
seq_1 <- seq(0.0, 10.0, by = 0.1)
seq_2 <- seq(10.0, 0.0, by = -0.1)
```

## Putting plots in a single figure {.smaller}

```{r, echo=TRUE, outwidth="50%", out.height="50%"}
par(mfrow=c(2,2))
plot (seq_1, xlab="time", ylab ="p in population 1", type = "p", col = 'red')
plot (seq_2, xlab="time", ylab ="p in population 2", type = "p", col = 'green')
plot (seq_square, xlab="time", ylab ="p2 in population 2", type = "p", col = 'blue')
plot (seq_square_new, xlab="time", ylab ="p in population 1", type = "l", col = 'yellow')
```


## Creating Data Frames in R {.smaller}

-   As you have seen, in R you can generate your own random data set drawn from nearly any distribution very easily.
-   Often we will want to use collected data.
-   Now, let’s make a dummy dataset to get used to dealing with data frames
-   Set up three variables (hydrogel_concentration, compression and conductivity) as vectors

```{r, echo=TRUE}
hydrogel_concentration <- factor(c("low", "high", "high", "high", "medium", "medium", "medium","low"))
compression <- c(3.4, 3.4, 8.4, 3, 5.6, 8.1, 8.3, 4.5)
conductivity <- c(0, 9.2, 3.8, 5, 5.6, 4.1, 7.1, 5.3)
```

-   Create a data frame where vectors become columns

```{r, echo=TRUE}
mydata <- data.frame(hydrogel_concentration, compression, conductivity)
row.names(mydata) <- c("Sample_1", "Sample_2", "Sample_3", "Sample_4", 
                       "Sample_5", "Sample_6", "Sample_7", "Sample_8")
```

-   Now you have a hand-made data frame with row names
-   Take a look at it in the data section of RStudio

## Reading in and Exporting Data Frames

\

```{r, eval = FALSE, echo = TRUE}
YourFile <- read.table('yourfile.csv', header=T, row.names=1, sep=',')
YourFile <- read.table('yourfile.txt', header=T, row.names=1, sep='\t')
```

\

```{r, eval = FALSE, echo = TRUE}
write.table(YourFile, "yourfile.csv", quote=F, row.names=T, sep=",")
write.table(YourFile, "yourfile.txt", quote=F, row.names=T, sep="\t")
```

## Indexing in data frames

-   Next up - indexing just a subset of the data
-   This is a very important idea in R, that you can analyze just a subset of the data.
-   This is analyzing only the data in the file you made that has the factor value 'mixed'.

```{r, echo=TRUE, eval=FALSE}
print (YourFile[,2])
print (YourFile$variable)
print (YourFile[2,])
plot (YourFile$variable1, YourFile$variable2)
```


## Practice reading in transcriptomic data {.smaller}

```{r, eval = FALSE, echo = TRUE}
RNAseq_Data <- read.table('<name_of_file>', header=TRUE, sep=',')

print (RNAseq_Data)
head (RNAseq_Data)
tail (RNAseq_Data)

print (RNAseq_Data[,2])
print (RNAseq_Data[1,])
print (RNAseq_Data[1,2])
print (RNAseq_Data$ENSGACG00000000010)
print (RNAseq_Data$ENSGACG00000000010>45.0)
```

## Summary stats and figures {.smaller}

```{r, eval = FALSE, echo = TRUE}
summary1 <- summary(RNAseq_Data $ENSGACG00000000003)
print (summary1)

hist(RNAseq_Data $ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003)
boxplot(RNAseq_Data$ENSGACG00000000003~RNAseq_Data$Population)
plot(RNAseq_Data $ENSGACG00000000003, RNAseq_Data$ENSGACG00000000003)

boxplot(RNAseq_Data $ENSGACG00000000003~RNAseq_Data$Treatment, 
        col = "red", ylab = "Expression Level", xlab = "Treatment level", 
        border ="orange", 
        main = "Boxplot of variation in gene expression across microbiota treatments")
```

# BREAK

------------------------------------------------------------------------

# Markdown and Quarto Markdown

## What is Markdown? {.smaller}

- a very light weight but flexible editing language
- allows the `rendering` of markdown text to a variety of different formats (pdf, html, web pages, powerpoint, etc....)
- very simple and common across science and engineering
- involves a `yaml` header (which stands for `yet another markdown language`)
- similar to `html` but simpler

## What is Quarto Markdown? {.smaller}

Quarto `(https://quarto.org/)` provides a next-generation authoring framework for data science.

-   `Quarto Markdown` is a powerful scientific and technical publishing system that extends the lightweight markup language Markdown for creating formatted text using a plain-text editor
-   You can use `Quarto Markdown` files to save and execute code \*and\* generate high quality reports that can be shared with an audience.
-   You can make html, pdf, docx, websites, slides, books, and more.

When you save your document, the output will be in the format indicated in your YAML header. This could be an html document, Word document, pdf, website, book, etc.

## Getting started {.smaller}

-   Open a new Quarto Markdown document from RStudio

-   You should see a document that is already populated with some example text and code

## This is Quarto Markdown {.smaller}

```{markdown, eval=F, echo=T}
---
title: "title"
output: html_document
---
```

- Note that `---` delineates the `yaml header`. 
- This controls document-level settings and rendering.
- There are a lot of options that you can put in the `yaml header`

## This is Quarto Markdown {.smaller}

```{markdown, eval=T, echo=T}
# Header size 1

## Header size 2

I can write text and it will look normal!

I can still run code!
```


## This is Quarto Markdown {.smaller}

```{r, eval=FALSE, echo=TRUE}
# ```{r code_chunk_name option1 option 2}
# x <- 5
# print(x^2)
# ```
```

-   By default, the code and output will print to your document
-   Note the back tick marks - these delineate the code
-   The {r} indicates this is r code

## Navigating chunks {.smaller}

-   Easy to run chunks - a few ways to do this
-   Top right of each chunk - just run, run all above, etc.
-   Cog wheel at the top of the Quarto Markdown document - run options
-   Keyboard shortcuts

## Chunk options {.smaller}

-   `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.
-   `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.
-   `message = FALSE` prevents messages that are generated by code from appearing in the finished file.
-   `warning = FALSE` prevents warnings that are generated by code from appearing in the finished.
-   `fig.cap = "..."` adds a caption to graphical results.
-   `eval = FALSE` shows code, but doesn't run it
-   `results = 'hide'` runs code, but doesn't show results
-   `fig.height, fig.width` set the dimensions of plots
-   `cache = TRUE` saves the results of a code chunk, can be helpful if a particular code chunk is intensive


## Quarto Markdown formatting {.smaller}

``` markdown
*italic*
**bold**
$A = \pi \times r^{2}$
```

*italic*

**bold**

$A = \pi \times r^{2}$

## Quarto Markdown lists {.smaller}

``` markdown
* Item 1
* Item 2
    + Item 2a
    + Item 2b
```

-   Item 1
-   Item 2
    -   Item 2a
    -   Item 2b

## Quarto Markdown lists {.smaller}

``` markdown
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b
```

1.  Item 1
2.  Item 2
3.  Item 3
    -   Item 3a
    -   Item 3b



## Typesetting Mathematics

-   Since 'tex' is embedded in Quarto Markdown you can typeset mathematics easily using LaTeX (https://en.wikibooks.org/wiki/LaTeX)

-   Essentially, just surround your mathematics in \$ symbols

-   \$\$ centers the expressions

## Typeset mathematics

Poisson Sampling Equation

```{r, eval=FALSE, echo=TRUE}
$$\large Pr(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$
```

$$\large Pr(Y=r) = \frac{e^{-\mu}\mu^r}{r!}$$

## Differential Equations {.smaller}

```{r, eval=FALSE, echo=TRUE}
$$\iint xy^2\,dx\,dy =\frac{1}{6}x^2y^3$$
```

$$\iint xy^2\,dx\,dy =\frac{1}{6}x^2y^3$$


## In-line versus fenced {.smaller}

```{r, eval=FALSE, echo=TRUE}
This equation, $y=\frac{1}{2}$, is included inline
```

This equation, $y=\frac{1}{2}$, is included inline

```{r, eval=FALSE, echo=TRUE}
Whereas this equation, $$y=\frac{1}{2}$$, is put on a separate line
```

Whereas this equation $$y=\frac{1}{2}$$ is put on a separate line

## Matrix formulations {.smaller}

```{r, eval=FALSE, echo=TRUE}
$$	\begin{matrix}
		-2 & 1 & 0 & 0 & \cdots & 0  \\
		1 & -2 & 1 & 0 & \cdots & 0  \\
		0 & 1 & -2 & 1 & \cdots & 0  \\
		0 & 0 & 1 & -2 & \ddots & \vdots \\
		\vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
		0 & 0 & 0 & \cdots & 1 & -2
	\end{matrix} $$
```

$$  \begin{matrix}
        -2 & 1 & 0 & 0 & \cdots & 0  \\
        1 & -2 & 1 & 0 & \cdots & 0  \\
        0 & 1 & -2 & 1 & \cdots & 0  \\
        0 & 0 & 1 & -2 & \ddots & \vdots \\
        \vdots & \vdots & \vdots & \ddots & \ddots & 1  \\
        0 & 0 & 0 & \cdots & 1 & -2
    \end{matrix} $$
    
## Quarto Markdown is very flexible {.smaller}

-   You can use Quarto templates and open as a new Quarto Markdown file
-   Better yet there are extensions that add functionality
    -   books
    -   journal articles
    -   slide shows (these slides!)
    -   interactive exercises
    
## More on Quarto Markdown {.smaller}

-   https://quarto.org/docs/authoring/
-   https://quarto.org/docs/reference/
-   Lots more fun stuff you can do!

------------------------------------------------------------------------

# `Vibe coding` with the ***help*** of Large Language Models (LLMs)

------------------------------------------------------------------------

# Git and GitHub

## What is Git? {.smaller}

-   **Git** is a distributed version control system
-   Tracks changes in your code over time
-   Allows collaboration with others
-   Maintains complete history of all changes
-   Essential tool for reproducible research
-   Works locally on your computer

## What is GitHub? {.smaller}

-   **GitHub** is a web-based hosting service for Git repositories
-   Provides cloud storage for your code
-   Enables collaboration features:
    -   Pull requests
    -   Issues tracking
    -   Project management
-   Hosts documentation and websites (GitHub Pages)
-   Free for public repositories

## Version Control Benefits {.smaller}

```{mermaid}
%%| fig-width: 10
%%| fig-height: 6
flowchart LR
    A[Working Directory] -->|git add| B[Staging Area]
    B -->|git commit| C[Local Repository]
    C -->|git push| D[Remote Repository]
    D -->|git pull| A
    
    style A fill:#f9d71c
    style B fill:#87ceeb
    style C fill:#98fb98
    style D fill:#ffa07a
```

-   **Track Changes**: Every modification is recorded
-   **Collaboration**: Multiple people can work simultaneously
-   **Backup**: Your code is safe in multiple places
-   **Experimentation**: Try new things without breaking working code

## Git Workflow Concept {.smaller}

```{mermaid}
%%| fig-width: 10
%%| fig-height: 7
graph TB
    subgraph "Your Computer"
        WD[Working Directory<br/>Your actual files]
        SA[Staging Area<br/>Files ready to commit]
        LR[Local Repository<br/>Committed changes]
    end
    
    subgraph "GitHub"
        RR[Remote Repository<br/>Shared with others]
    end
    
    WD -->|git add| SA
    SA -->|git commit| LR
    LR -->|git push| RR
    RR -->|git pull| WD
    
    style WD fill:#e1f5fe
    style SA fill:#fff3e0
    style LR fill:#f3e5f5
    style RR fill:#e8f5e9
```

## Setting Up Git {.smaller}

### 1. Install Git
``` bash
# Check if Git is installed
git --version

# Ubuntu/Debian
sudo apt-get install git

# macOS (with Homebrew)
brew install git

# Windows: Download from git-scm.com
```

### 2. Configure Git
``` bash
# Set your name and email
git config --global user.name "Your Name"
git config --global user.email "your.email@example.com"

# Check your settings
git config --list
```
# GitHub

## GitHub for Scientific Collaboration {.smaller}

### Perfect for Research Projects:

-   **Reproducibility**: Complete history of analysis development
-   **Collaboration**: Work with colleagues worldwide
-   **Documentation**: README files, wikis, and GitHub Pages
-   **Issues**: Track bugs, ideas, and to-dos
-   **Releases**: Tag specific versions for publications
-   **DOI Integration**: Link with Zenodo for citations

## Resources and Help {.smaller}

### Learning Resources:
-   GitHub Docs `(https://docs.github.com)`
-   Pro Git Book `(https://git-scm.com/book) (free online)`
-   GitHub Learning Lab `(https://github.com/apps/github-learning-lab)`
-   Quarto + GitHub Pages `(https://quarto.org/docs/publishing/github-pages.html)`

### Getting Help:
``` bash
# Git help
git help <command>
git <command> --help

# Quick reference
git help everyday
```

## Creating a GitHub Account {.smaller}

### Steps to Get Started:

1. Go to [github.com](https://github.com)
2. Click "Sign up"
3. Choose a username (professional, memorable)
4. Add your university email
5. Verify your account
6. Consider GitHub Student Developer Pack:
   - Free GitHub Pro while you're a student
   - Access to development tools
   - Apply at [education.github.com](https://education.github.com)

## Basic Git Commands {.smaller}

### Initialize a Repository
``` bash
# Create a new repository
git init

# Clone an existing repository
git clone https://github.com/username/repository.git
```

### Track Changes
``` bash
# Check status
git status

# Add files to staging
git add filename.txt        # Add specific file
git add .                   # Add all changes

# Commit changes
git commit -m "Descriptive message about changes"
```

## Working with Remotes {.smaller}

``` bash
# Add a remote repository
git remote add origin https://github.com/username/repo.git

# View remotes
git remote -v

# Push changes to GitHub
git push origin main

# Pull changes from GitHub
git pull origin main

# Fetch without merging
git fetch origin
```

## Branching and Merging {.smaller}

```{mermaid}
%%| fig-width: 10
%%| fig-height: 6
gitGraph
    commit id: "Initial commit"
    commit id: "Add README"
    branch feature
    checkout feature
    commit id: "New feature"
    commit id: "Fix bug"
    checkout main
    commit id: "Update docs"
    merge feature
    commit id: "Continue work"
```

``` bash
# Create and switch to new branch
git checkout -b feature-branch

# Switch branches
git checkout main

# Merge branch
git merge feature-branch
```

## Common Git Workflow {.smaller}

### Daily Development Cycle

``` bash
# 1. Start your day - update your local copy
git pull origin main

# 2. Create a feature branch
git checkout -b new-feature

# 3. Make changes and commit regularly
git add .
git commit -m "Add new analysis function"

# 4. Push your branch
git push origin new-feature

# 5. Create pull request on GitHub (via website)
```

## Viewing History {.smaller}

``` bash
# View commit history
git log

# Prettier one-line format
git log --oneline

# With graph
git log --graph --oneline --all

# Show changes in commits
git log -p

# Show last n commits
git log -3
```

## GitHub Pages {.smaller}

### Host Your Research Website for Free!

```{mermaid}
%%| fig-width: 10
%%| fig-height: 5
flowchart LR
    A[Quarto Documents] -->|Render| B[HTML Files]
    B -->|git push| C[GitHub Repository]
    C -->|Automatic| D[GitHub Pages Website]
    D --> E[yourname.github.io/project]
    
    style A fill:#e3f2fd
    style E fill:#c8e6c9
```

### Setup Steps:
1. Create repository
2. Push your HTML files
3. Enable Pages in repository settings
4. Your site is live at `https://username.github.io`

## GitHub Pages Configuration {.smaller}

### In your repository:

1. Go to Settings → Pages
2. Source: Deploy from a branch
3. Branch: main (or gh-pages)
4. Folder: / (root) or /docs

### For Quarto projects:
``` yaml
# In _quarto.yml
project:
  type: website
  output-dir: docs

publish:
  gh-pages:
    branch: gh-pages
```

``` bash
# Publish with Quarto
quarto publish gh-pages
```

## Best Practices {.smaller}

### Commit Messages
-   Be descriptive and concise
-   Use present tense: "Add feature" not "Added feature"
-   Reference issues: "Fix bug #123"

### Repository Organization
``` 
project/
├── README.md           # Project description
├── data/              # Raw data (consider .gitignore)
├── analysis/          # Analysis scripts
├── results/           # Output files
└── docs/              # Documentation
```

## .gitignore File {.smaller}

### Exclude files from version control:

``` bash
# Create .gitignore file
touch .gitignore
```

### Common patterns to exclude:

``` bash
# Data files
*.fastq
*.fasta
*.bam
*.vcf
data/raw/*

# R files
.Rhistory
.RData
.Rproj.user

# Output files
results/temp/*
*.log

# System files
.DS_Store
```

